Jane, John … Leslie? A Historical Method for Algorithmic Gender Detection

### 2014-09-15 article.Rmd
Citations

But researchers studying a longer timespan need to take into account the changing nature of naming practices. If that same education researcher wanted to compare the demographics of contemporary urban schools with those from the early twentieth century, contemporary tools like Genderize.io would misidentify Leslie Fay's gender in historical datasets. The "Leslie" problem is not just for researchers who think of themselves as historians. Anyone studying a period longer than a few years, or anyone studying a group whose demographics do not match those of a contemporary tool like Genderize.io will also encounter this problem. As of 2012, the average American has a lifespan of nearly seventy-nine years–--more than enough time for naming practices to change quite dramatically between their birth and death [World Bank 2014]. Predicting gender from first names therefore requires a fundamentally historical method.

@@

The Social Security Administration (SSA) "[Baby Names](http://catalog.data.gov/dataset/baby-names-from-social-security-card-applications-national-level-data)" dataset was created as a result of the Social Security Act of 1935 during the New Deal. ^[The Social Security Administration also provides a related dataset which contains the names of applicants broken down by state and year since 1910 \[Social Security Administration 2014-b]\.] This dataset contains a list of first names with the number of male and female uses of the name in every year beginning with 1880 and ending with 2012. The SSA list includes any name which was used more than five times in a given year, thereby capturing all but the most infrequently used names from each year. The description "baby names" provided by the SSA is a serious misnomer. When Social Security first became available during the New Deal, its first beneficiaries were adults past or near retirment age. The dataset goes back to 1880, the birth year for a 55 year-old adult when Social Security was enacted. Even after 1935, registration at birth for Social Security was not mandatory until 1986. As we will demonstrate below, the way in which the data was gathered requires an adjustment to our predictions of gender.^[Shane Landrum has studie dthe history of birth registration in the United States \[Landrum 2014]\.  It is well-known among American historians that the Social Security Act was written to exclude farmworkers as a means of also excluding African Americans. In examining the dataset we have not found any reason to believe that this policy noticably skewed the prediction of gender from names.]

The IPUMS-USA dataset, contributed by Benjamin Schmidt, contains records from the United States decennial census from 1790 to 1930. This dataset contains the birth year and numbers of males and females under the age of 62 for all the years in that range. This data has been aggregated by IPUMS at the University of Minnesota and is released as a sample of the total census data. Unlike the SSA dataset, which includes 100% of every name reported to the Social Security Administration, the IPUMS data contains 5% or 10% samples. Because the `gender()` function relies on proportions of uses of names, rather than raw counts of people with the names, the sampling does not diminish the reliability of the function's predicitons.^[Steven Ruggles, J. Trent Alexander, Katie Genadek, Ronald Goeken, Matthew B. Schroeder, and Matthew Sobek. Integrated Public Use Microdata Series: Version 5.0 [Machine-readable database]. Minneapolis, MN: Minnesota Population Center [producer and distributor], 2010.]

These two datasets (and any future datasets to be added to the package) thus contain a simple table of data. The columns

^[The format of this data departs from the more generally desirable "tidy data" framework proposed by Hadley Wickham for performance reasons. By keeping the data in a "wider" format instead of a tidy format as defined by Wickham, the function is able to make its predictions faster. Hadley Wickham, "[Tidy Data](http://vita.had.co.nz/papers/tidy-data.html)," *Journal of Statistical Software* (forthcoming).] The columns contain the `name`, `year`, and number of `female` and `male` uses of a name. Each row contains a combination for a name and a year.

@@

@@

>>>>

But researchers studying a longer timespan need to take into account the changing nature of naming practices. If that same education researcher wanted to compare the demographics of contemporary urban schools with those from the early twentieth century, contemporary tools like Genderize.io would misidentify Leslie Fay's gender in historical datasets. The "Leslie" problem is not just for researchers who think of themselves as historians. Anyone studying a period longer than a few years, or anyone studying a group whose demographics do not match those of a contemporary tool like Genderize.io will also encounter this problem. As of 2012, the average American has a lifespan of nearly seventy-nine years–--more than enough time for naming practices to change quite dramatically between their birth and death \[World Bank 2014]\. Predicting gender from first names therefore requires a fundamentally historical method.

@@

The Social Security Administration (SSA) "[Baby Names](http://catalog.data.gov/dataset/baby-names-from-social-security-card-applications-national-level-data)" dataset was created as a result of the Social Security Act of 1935 during the New Deal. ^[The Social Security Administration also provides a related dataset which contains the names of applicants broken down by state and year since 1910 \[Social Security Administration 2014-b]\.] This dataset contains a list of first names with the number of male and female uses of the name in every year beginning with 1880 and ending with 2012. The SSA list includes any name which was used more than five times in a given year, thereby capturing all but the most infrequently used names from each year. The description "baby names" provided by the SSA is a serious misnomer. When Social Security first became available during the New Deal, its first beneficiaries were adults past or near retirment age. The dataset goes back to 1880, the birth year for a 55 year-old adult when Social Security was enacted. Even after 1935, registration at birth for Social Security was not mandatory until 1986. As we will demonstrate below, the way in which the data was gathered requires an adjustment to our predictions of gender.^[Shane Landrum has studied the history of birth registration in the United States \[Landrum 2014]\.  It is well-known among American historians that the Social Security Act was written to exclude farmworkers as a means of also excluding African Americans. In examining the dataset we have not found any reason to believe that this policy noticably skewed the prediction of gender from names.]

The IPUMS-USA dataset, contributed by Benjamin Schmidt, contains records from the United States decennial census from 1790 to 1930. This dataset contains the birth year and numbers of males and females under the age of 62 for all the years in that range. This data has been aggregated by IPUMS at the University of Minnesota and is released as a sample of the total census data. Unlike the SSA dataset, which includes 100% of every name reported to the Social Security Administration, the IPUMS data contains 5% or 10% samples. Because the `gender()` function relies on proportions of uses of names, rather than raw counts of people with the names, the sampling does not diminish the reliability of the function's predicitons \[Ruggles et al. 2010]\.

These two datasets (and any future datasets to be added to the package) thus contain a simple table of data. ^[The format of this data departs from the more generally desirable "tidy data" framework proposed by Hadley Wickham for performance reasons. By keeping the data in a "wider" format instead of a tidy format as defined by Wickham, the function is able to make its predictions faster \[Wickham forthcoming]\.] The columns contain the `name`, `year`, and number of `female` and `male` uses of a name. Each row contains a combination for a name and a year.

@@

Ruggles et al. 2010. Ruggles, Steven, J. Trent Alexander, Katie Genadek, Ronald Goeken, Matthew B. Schroeder, and Matthew Sobek. Integrated Public Use Microdata Series: Version 5.0 \[Machine-readable database]\. Minneapolis, MN: Minnesota Population Center \[producer and distributor]\, 2010.

@@

Wickham, forthcoming. Hadley Wickham, "Tidy Data." *Journal of Statistical Software* (forthcoming), http://vita.had.co.nz/papers/tidy-data.html.
